### 2018-01-23 submission/manuscript.Rmd
add changes to results and discussion based on updated selected models

For the genera-based models both tissue and stool studies had a median AUC decrease in model classification when only CRC-associated genera versus all genera were included [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both all and CRC-associated genera models had a similar ability to detect adenomas or carcinomas [Figure S4-S6]. The OTU-based models showed a similar trend to the genera models where a generalized decrease in median AUC was observed when only OTUs from the CRC-associated genera were used versus the full community of OTUs for both adenoma and carcinoma, with the largest median change for stool in carcinoma models and for tissue in adenoma models [Figure 4 & 5].

@@

The data presented herein supports the importance of mouth-associated genera for carcinoma but not necessarily adenoma formation. A proposed mechanism by which these mouth-associated microbes cause CRC stems from the creation of a vicious cycle involving host inflammation and repair. Both the drastically increased carcinoma RR of mouth-associated genera versus $\alpha$-diversity metrics and increasing RR with more mouth-associated genera positivity [Figure 3C], are highly supportive of their importance. However, our observations show that when adding the community context to our Random Forest models along with the mouth-associated genera, the model AUC increases [Figure 4 & 5]. This suggests the resident community cannot be ignored completely and needs to be accounted for when investigating mouth-associated microbes and carcinoma. Conversely, using the present data, it is less likely that mouth-associated microbes are correlated with adenoma development. The step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. Yet, this change in diversity, at this early stage of disease, could be focal to the adenoma itself. The poor performance of the Random Forest models for classifying adenoma, based only on the stool microbiota, would suggest that this is the case. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

>>>>

For the genera-based models the training set median AUC for model classification was similar for select genera and full genera models, for both tissue and stool studies [Figure S2-S3]. With respect to the test sets, comprised of genera data from other studies, both all and select genera models had a similar ability to detect adenomas or carcinomas, with the select genera models performing better in some instances [Figure S4-S6]. Conversely, the OTU-based models showed a slight decrease in median AUC between the full and select models, with one exception being carcinoma models for matched tissue [Figure 4 & 5].

@@

The data presented herein supports the importance of select genera for carcinoma but not necessarily adenoma formation. Our observations show that when only using the select genera to create models for carcinoma the AUC are similar to using all genera or OTUs [Figure 4 & 5]. This suggests that an interplay between a select number of potentially protective and exacerbating microbes, within the GI community, are crucial for carcinoma formation. Importantly, it suggests that there may be key members of the community that might be studied further to potentially reduce the risk of carcinoma. Conversely, using the present data, it is clear that new approaches may need to be used to identify members of the community that are associated with adenoma. Regardless of sample type and whether a full or select model was used our Random Forest models consistently performed poorly. Yet, the step-wise decrease in diversity suggests that the adenoma community is not normal but has changed subtly [Figure 1]. This change in diversity, at this early stage of disease, could be focal to the adenoma itself. One potential hypothesis from these observations is that at early stages of the disease, how the host interacts with these subtle changes is what ultimately leads to a thoroughly dysfunctional community that is supportive of CRC genesis.

### 2018-01-23 submission/manuscript.tex
add changes to results and discussion based on updated selected models

For the genera-based models both tissue and stool studies had a median
AUC decrease in model classification when only CRC-associated genera
versus all genera were included {[}Figure S2-S3{]}. With respect to the
test sets, comprised of genera data from other studies, both all and
CRC-associated genera models had a similar ability to detect adenomas or
carcinomas {[}Figure S4-S6{]}. The OTU-based models showed a similar
trend to the genera models where a generalized decrease in median AUC
was observed when only OTUs from the CRC-associated genera were used
versus the full community of OTUs for both adenoma and carcinoma, with
the largest median change for stool in carcinoma models and for tissue
in adenoma models {[}Figure 4 \& 5{]}.

@@

The data presented herein supports the importance of mouth-associated
genera for carcinoma but not necessarily adenoma formation. A proposed
mechanism by which these mouth-associated microbes cause CRC stems from
the creation of a vicious cycle involving host inflammation and repair.
Both the drastically increased carcinoma RR of mouth-associated genera
versus \(\alpha\)-diversity metrics and increasing RR with more
mouth-associated genera positivity {[}Figure 3C{]}, are highly
supportive of their importance. However, our observations show that when
adding the community context to our Random Forest models along with the
mouth-associated genera, the model AUC increases {[}Figure 4 \& 5{]}.
This suggests the resident community cannot be ignored completely and
needs to be accounted for when investigating mouth-associated microbes
and carcinoma. Conversely, using the present data, it is less likely
that mouth-associated microbes are correlated with adenoma development.
The step-wise decrease in diversity suggests that the adenoma community
is not normal but has changed subtly {[}Figure 1{]}. Yet, this change in
diversity, at this early stage of disease, could be focal to the adenoma
itself. The poor performance of the Random Forest models for classifying
adenoma, based only on the stool microbiota, would suggest that this is
the case. One potential hypothesis from these observations is that at
early stages of the disease, how the host interacts with these subtle
changes is what ultimately leads to a thoroughly dysfunctional community
that is supportive of CRC genesis.

>>>>

For the genera-based models the training set median AUC for model
classification was similar for select genera and full genera models, for
both tissue and stool studies {[}Figure S2-S3{]}. With respect to the
test sets, comprised of genera data from other studies, both all and
select genera models had a similar ability to detect adenomas or
carcinomas, with the select genera models performing better in some
instances {[}Figure S4-S6{]}. Conversely, the OTU-based models showed a
slight decrease in median AUC between the full and select models, with
one exception being carcinoma models for matched tissue {[}Figure 4 \&
5{]}.

@@

The data presented herein supports the importance of select genera for
carcinoma but not necessarily adenoma formation. Our observations show
that when only using the select genera to create models for carcinoma
the AUC are similar to using all genera or OTUs {[}Figure 4 \& 5{]}.
This suggests that an interplay between a select number of potentially
protective and exacerbating microbes, within the GI community, are
crucial for carcinoma formation. Importantly, it suggests that there may
be key members of the community that might be studied further to
potentially reduce the risk of carcinoma. Conversely, using the present
data, it is clear that new approaches may need to be used to identify
members of the community that are associated with adenoma. Regardless of
sample type and whether a full or select model was used our Random
Forest models consistently performed poorly. Yet, the step-wise decrease
in diversity suggests that the adenoma community is not normal but has
changed subtly {[}Figure 1{]}. This change in diversity, at this early
stage of disease, could be focal to the adenoma itself. One potential
hypothesis from these observations is that at early stages of the
disease, how the host interacts with these subtle changes is what
ultimately leads to a thoroughly dysfunctional community that is
supportive of CRC genesis.
